---
title: "bnlearn Simulations"
author: "Tim Vigers"
date: "10/23/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,message = FALSE)
knitr::opts_knit$set(root.dir = "/home/tim/Documents/GitHub/MS-Thesis")
library(bnlearn)
library(DiagrammeR)
library(knitr)
source("./code/networks/network_structures.R")
set.seed(1017)
```

```{r parameters}
nsim = 10000
n = 160
## True values
## T1D -> metab
alpha <- 2 # Slope
## methyl | T1D, metab
beta <- 8 # T1D effect
gamma <- 5 # Metab effect
```

```{r check functions}
sim_eval <- function(sims,correct,str_number){
  checks = lapply(sims,function(x){
    # Add noise to T1D
    x$t1d = as.numeric(x$t1d) + rnorm(n,sd = 0.0001)
    learned <- hc(x)
    mod <- modelstring(learned)
    check <- ifelse(mod == correct,1,0)
    l <- score(learned,x)
    c <- score(model2network(correct),x)
    return(c(check,l,c,mod))
  })
  res <- as.data.frame(do.call(rbind,checks))
  res <- cbind(str_number,res)
  colnames(res) = c("str_number","correct","bic_learned","bic_correct",
                    "learned_struct")
  res[,1:4] <- lapply(res[,1:4],as.numeric)
  res$bic_diff <- res$bic_correct - res$bic_learned
  return(res[,c("str_number","correct","bic_learned","bic_correct","bic_diff",
                "learned_struct")])
}
table_eval <- function(eval_res,correct){
  plot(model2network(correct),main = "Correct Structure")
  sum_t = as.data.frame(table(eval_res$learned_struct))
  colnames(sum_t)[1] = "Structure"
  print(kable(sum_t))
  apply(sum_t,1,function(r){
    plot(model2network(as.character(r["Structure"])),
         main = paste0("n = ",r["Freq"]))
  })
}
plot_eval <- function(eval_res){
  hist(eval_res$bic_learned,xlab = "Learned BIC",main = "")
  hist(eval_res$bic_learned,xlab = "Correct BIC",main = "")
  hist(eval_res$bic_diff,xlab = "BIC Correct - BIC Learned",main = "")
}
```

# Structure 1

```{r}
correct = "[t1d][metab|t1d][methyl|t1d:metab]"
correct = modelstring(model2network(correct))
# Simulate
dfs <- lapply(1:nsim,function(x){
  set.seed(1016 + x)
  # Simulate data
  t1d <- rbinom(n,1,prob = 0.5)
  metab <- rnorm(n, mean = alpha*t1d) 
  methyl <- rnorm(n, mean = beta*t1d + gamma*metab)
  df <- data.frame(t1d = t1d,metab = metab,methyl = methyl)
  return(df)
})
str_1_res <- sim_eval(dfs,correct,1)
```

Out of `r nsim` simulations, bnlearn picked the correct structure `r sum(str_1_res$correct)` times (`r sum(str_1_res$correct)/nrow(str_1_res) * 100`%).

```{r,results='asis'}
invisible(table_eval(str_1_res,correct))
```

```{r}
plot_eval(str_1_res)
```

# Structure 2

```{r}
correct = "[t1d][methyl|t1d][metab|methyl:t1d]"
correct = modelstring(model2network(correct))
# Simulate
dfs <- lapply(1:nsim,function(x){
  set.seed(1016 + x)
  # Simulate data
  t1d <- rbinom(n,1,prob = 0.5)
  methyl <- rnorm(n, mean = beta*t1d)
  metab <- rnorm(n, mean = alpha*t1d + gamma*methyl) 
  df <- data.frame(t1d = t1d,metab = metab,methyl = methyl)
  return(df)
})
str_2_res <- sim_eval(dfs,correct,2)
```

Out of `r nsim` simulations, bnlearn picked the correct structure `r sum(str_2_res$correct)` times (`r sum(str_2_res$correct)/nsim * 100`%).

```{r,results='asis'}
invisible(table_eval(str_2_res,correct))
```

```{r}
plot_eval(str_2_res)
```

# Structure 3

```{r}
correct = "[methyl][metab|methyl][t1d|metab:methyl]"
correct = modelstring(model2network(correct))
# Simulate
dfs <- lapply(1:nsim,function(x){
  set.seed(1016 + x)
  # Simulate data
  methyl <- rnorm(n,5)
  metab <- rnorm(n, mean = gamma*methyl)
  t1d <- rbinom(n,1,prob = plogis(alpha*metab + beta*methyl))
  df <- data.frame(t1d = t1d,metab = metab,methyl = methyl)
  return(df)
})
str_3_res <- sim_eval(dfs,correct,3)
```

Out of `r nsim` simulations, bnlearn picked the correct structure `r sum(str_3_res$correct)` times (`r sum(str_3_res$correct)/nsim * 100`%).

```{r,results='asis'}
invisible(table_eval(str_3_res,correct))
```

```{r}
plot_eval(str_3_res)
```

# Structure 4

```{r}
correct = "[methyl][metab|methyl:t1d][t1d|methyl]"
correct = modelstring(model2network(correct))
# Simulate
dfs <- lapply(1:nsim,function(x){
  set.seed(1016 + x)
  # Simulate data
  methyl <- rnorm(n,5)
  t1d <- rbinom(n,1,prob = plogis(beta*methyl))
  metab <- rnorm(n, mean = gamma*methyl + alpha*t1d)
  df <- data.frame(t1d = t1d,metab = metab,methyl = methyl)
  return(df)
})
str_4_res <- sim_eval(dfs,correct,4)
```

Out of `r nsim` simulations, bnlearn picked the correct structure `r sum(str_4_res$correct)` times (`r sum(str_4_res$correct)/nsim * 100`%).

```{r,results='asis'}
invisible(table_eval(str_4_res,correct))
```

```{r}
plot_eval(str_4_res)
```

# Structure 5

```{r}
correct = "[metab][methyl|metab][t1d|metab:methyl]"
correct = modelstring(model2network(correct))
# Simulate
dfs <- lapply(1:nsim,function(x){
  set.seed(1016 + x)
  # Simulate data
  metab <- rnorm(n,5)
  methyl <- rnorm(n, mean = gamma*metab)
  t1d <- rbinom(n,1,prob = plogis(alpha*metab + beta*methyl))
  df <- data.frame(t1d = t1d,metab = metab,methyl = methyl)
  return(df)
})
str_5_res <- sim_eval(dfs,correct,5)
```

Out of `r nsim` simulations, bnlearn picked the correct structure `r sum(str_5_res$correct)` times (`r sum(str_5_res$correct)/nsim * 100`%).

```{r,results='asis'}
invisible(table_eval(str_5_res,correct))
```

```{r}
plot_eval(str_5_res)
```

# Structure 6

```{r}
correct = "[metab][t1d|metab][methyl|metab:t1d]"
correct = modelstring(model2network(correct))
# Simulate
dfs <- lapply(1:nsim,function(x){
  set.seed(1016 + x)
  # Simulate data
  metab <- rnorm(n,5)
  t1d <- rbinom(n,1,prob = plogis(alpha*metab))
  methyl <- rnorm(n, mean = gamma*metab + beta*t1d)
  df <- data.frame(t1d = t1d,metab = metab,methyl = methyl)
  return(df)
})
str_6_res <- sim_eval(dfs,correct,6)
```

Out of `r nsim` simulations, bnlearn picked the correct structure `r sum(str_6_res$correct)` times (`r sum(str_6_res$correct)/nsim * 100`%).

```{r,results='asis'}
invisible(table_eval(str_6_res,correct))
```

```{r}
plot_eval(str_6_res)
```

# Structure 7

```{r}
correct = "[t1d][metab|t1d][methyl|t1d]"
correct = modelstring(model2network(correct))
# Simulate
dfs <- lapply(1:nsim,function(x){
  set.seed(1016 + x)
  # Simulate data
  t1d <- rbinom(n,1,prob = 0.5)
  methyl <- rnorm(n, mean = beta*t1d)
  metab <- rnorm(n, mean = alpha*t1d) 
  df <- data.frame(t1d = t1d,metab = metab,methyl = methyl)
  return(df)
})
str_7_res <- sim_eval(dfs,correct,7)
```

Out of `r nsim` simulations, bnlearn picked the correct structure `r sum(str_7_res$correct)` times (`r sum(str_7_res$correct)/nsim * 100`%).

```{r,results='asis'}
invisible(table_eval(str_7_res,correct))
```

```{r}
plot_eval(str_7_res)
```

# Structure 8

```{r}
correct = "[methyl][t1d|methyl][metab|t1d]"
correct = modelstring(model2network(correct))
# Simulate
dfs <- lapply(1:nsim,function(x){
  set.seed(1016 + x)
  # Simulate data
  methyl <- rnorm(n,5)
  t1d <- rbinom(n,1,prob = plogis(beta*methyl))
  metab <- rnorm(n, mean = alpha*t1d) 
  df <- data.frame(t1d = t1d,metab = metab,methyl = methyl)
  return(df)
})
str_8_res <- sim_eval(dfs,correct,8)
```

Out of `r nsim` simulations, bnlearn picked the correct structure `r sum(str_8_res$correct)` times (`r sum(str_8_res$correct)/nsim * 100`%).

```{r,results='asis'}
invisible(table_eval(str_8_res,correct))
```

```{r}
plot_eval(str_8_res)
```

# Structure 9

```{r}
correct = "[metab][t1d|metab][methyl|t1d]"
correct = modelstring(model2network(correct))
# Simulate
dfs <- lapply(1:nsim,function(x){
  set.seed(1016 + x)
  # Simulate data
  metab <- rnorm(n,5)
  t1d <- rbinom(n,1,prob = plogis(alpha*metab))
  methyl <- rnorm(n, mean = beta*t1d) 
  df <- data.frame(t1d = t1d,metab = metab,methyl = methyl)
  return(df)
})
str_9_res <- sim_eval(dfs,correct,9)
```

Out of `r nsim` simulations, bnlearn picked the correct structure `r sum(str_9_res$correct)` times (`r sum(str_9_res$correct)/nsim * 100`%).

```{r,results='asis'}
invisible(table_eval(str_9_res,correct))
```

```{r}
plot_eval(str_9_res)
```

# Structure 10

```{r}
correct = "[metab][methyl][t1d|metab:methyl]"
correct = modelstring(model2network(correct))
# Simulate
dfs <- lapply(1:nsim,function(x){
  set.seed(1016 + x)
  # Simulate data
  metab <- rnorm(n,5)
  methyl <- rnorm(n,5)
  t1d <- rbinom(n,1,prob = plogis(alpha*metab + beta*methyl))
  df <- data.frame(t1d = t1d,metab = metab,methyl = methyl)
  return(df)
})
str_10_res <- sim_eval(dfs,correct,10)
```

Out of `r nsim` simulations, bnlearn picked the correct structure `r sum(str_10_res$correct)` times (`r sum(str_10_res$correct)/nsim * 100`%).

```{r,results='asis'}
invisible(table_eval(str_10_res,correct))
```

```{r}
plot_eval(str_10_res)
```

# Structure 11

```{r}
correct = "[methyl][metab|methyl][t1d|methyl]"
correct = modelstring(model2network(correct))
# Simulate
dfs <- lapply(1:nsim,function(x){
  set.seed(1016 + x)
  # Simulate data
  methyl <- rnorm(n,5)
  t1d <- rbinom(n,1,prob = plogis(beta*methyl))
  metab <- rnorm(n,mean = gamma*methyl)
  df <- data.frame(t1d = t1d,metab = metab,methyl = methyl)
  return(df)
})
str_11_res <- sim_eval(dfs,correct,11)
```

Out of `r nsim` simulations, bnlearn picked the correct structure `r sum(str_11_res$correct)` times (`r sum(str_11_res$correct)/nsim * 100`%).

```{r,results='asis'}
invisible(table_eval(str_11_res,correct))
```

```{r}
plot_eval(str_11_res)
```

# Structure 12

```{r}
correct = "[t1d][methyl|t1d][metab|methyl]"
correct = modelstring(model2network(correct))
# Simulate
dfs <- lapply(1:nsim,function(x){
  set.seed(1016 + x)
  # Simulate data
  t1d <- rbinom(n,1,0.5)
  methyl <- rnorm(n,mean = beta*t1d)
  metab <- rnorm(n,mean = gamma*methyl)
  df <- data.frame(t1d = t1d,metab = metab,methyl = methyl)
  return(df)
})
str_12_res <- sim_eval(dfs,correct,12)
```

Out of `r nsim` simulations, bnlearn picked the correct structure `r sum(str_12_res$correct)` times (`r sum(str_12_res$correct)/nsim * 100`%).

```{r,results='asis'}
invisible(table_eval(str_12_res,correct))
```

```{r}
plot_eval(str_12_res)
```

# Structure 13

```{r}
correct = "[metab][methyl|metab][t1d|methyl]"
correct = modelstring(model2network(correct))
# Simulate
dfs <- lapply(1:nsim,function(x){
  set.seed(1016 + x)
  # Simulate data
  metab <- rnorm(n,5)
  methyl <- rnorm(n,mean = gamma*metab)
  t1d <- rbinom(n,1,prob = plogis(beta*methyl))
  df <- data.frame(t1d = t1d,metab = metab,methyl = methyl)
  return(df)
})
str_13_res <- sim_eval(dfs,correct,13)
```

Out of `r nsim` simulations, bnlearn picked the correct structure `r sum(str_13_res$correct)` times (`r sum(str_13_res$correct)/nsim * 100`%).

```{r,results='asis'}
invisible(table_eval(str_13_res,correct))
```

```{r}
plot_eval(str_13_res)
```

# Structure 14

```{r}
correct = "[metab][t1d][methyl|metab:t1d]"
correct = modelstring(model2network(correct))
# Simulate
dfs <- lapply(1:nsim,function(x){
  set.seed(1016 + x)
  # Simulate data
  t1d <- rbinom(n,1,0.5)
  metab <- rnorm(n,5)
  methyl <- rnorm(n,mean = gamma*metab + beta*t1d)
  df <- data.frame(t1d = t1d,metab = metab,methyl = methyl)
  return(df)
})
str_14_res <- sim_eval(dfs,correct,14)
```

Out of `r nsim` simulations, bnlearn picked the correct structure `r sum(str_14_res$correct)` times (`r sum(str_14_res$correct)/nsim * 100`%).

```{r,results='asis'}
invisible(table_eval(str_14_res,correct))
```

```{r}
plot_eval(str_14_res)
```

# Structure 15

```{r}
correct = "[metab][methyl|metab][t1d|metab]"
correct = modelstring(model2network(correct))
# Simulate
dfs <- lapply(1:nsim,function(x){
  set.seed(1016 + x)
  # Simulate data
  metab <- rnorm(n,5)
  t1d <- rbinom(n,1,prob = plogis(alpha*metab))
  methyl <- rnorm(n,mean = gamma*metab)
  df <- data.frame(t1d = t1d,metab = metab,methyl = methyl)
  return(df)
})
str_15_res <- sim_eval(dfs,correct,15)
```

Out of `r nsim` simulations, bnlearn picked the correct structure `r sum(str_15_res$correct)` times (`r sum(str_15_res$correct)/nsim * 100`%).

```{r,results='asis'}
invisible(table_eval(str_15_res,correct))
```

```{r}
plot_eval(str_15_res)
```

# Structure 16

```{r}
correct = "[t1d][metab|t1d][methyl|metab]"
correct = modelstring(model2network(correct))
# Simulate
dfs <- lapply(1:nsim,function(x){
  set.seed(1016 + x)
  # Simulate data
  t1d <- rbinom(n,1,0.5)
  metab <- rnorm(n,mean = alpha*t1d)
  methyl <- rnorm(n,mean = gamma*metab)
  df <- data.frame(t1d = t1d,metab = metab,methyl = methyl)
  return(df)
})
str_16_res <- sim_eval(dfs,correct,16)
```

Out of `r nsim` simulations, bnlearn picked the correct structure `r sum(str_16_res$correct)` times (`r sum(str_16_res$correct)/nsim * 100`%).

```{r,results='asis'}
invisible(table_eval(str_16_res,correct))
```

```{r}
plot_eval(str_16_res)
```

# Structure 17

```{r}
correct = "[methyl][metab|methyl][t1d|metab]"
correct = modelstring(model2network(correct))
# Simulate
dfs <- lapply(1:nsim,function(x){
  set.seed(1016 + x)
  # Simulate data
  methyl <- rnorm(n,5)
  metab <- rnorm(n,mean = gamma*methyl)
  t1d <- rbinom(n,1,prob = plogis(alpha*metab))
  df <- data.frame(t1d = t1d,metab = metab,methyl = methyl)
  return(df)
})
str_17_res <- sim_eval(dfs,correct,17)
```

Out of `r nsim` simulations, bnlearn picked the correct structure `r sum(str_17_res$correct)` times (`r sum(str_17_res$correct)/nsim * 100`%).

```{r,results='asis'}
invisible(table_eval(str_17_res,correct))
```

```{r}
plot_eval(str_17_res)
```

# Structure 18

```{r}
correct = "[methyl][t1d][metab|methyl:t1d]"
correct = modelstring(model2network(correct))
# Simulate
dfs <- lapply(1:nsim,function(x){
  set.seed(1016 + x)
  # Simulate data
  t1d <- rbinom(n,1,0.5)
  methyl <- rnorm(n,5)
  metab <- rnorm(n,mean = gamma*methyl + alpha*t1d)
  df <- data.frame(t1d = t1d,metab = metab,methyl = methyl)
  return(df)
})
str_18_res <- sim_eval(dfs,correct,18)
```

Out of `r nsim` simulations, bnlearn picked the correct structure `r sum(str_18_res$correct)` times (`r sum(str_18_res$correct)/nsim * 100`%).

```{r,results='asis'}
invisible(table_eval(str_18_res,correct))
```

```{r}
plot_eval(str_18_res)
```

# Structure 19

```{r}
correct = "[methyl|t1d][t1d][metab]"
correct = modelstring(model2network(correct))
# Simulate
dfs <- lapply(1:nsim,function(x){
  set.seed(1016 + x)
  # Simulate data
  t1d <- rbinom(n,1,0.5)
  methyl <- rnorm(n,mean = beta*t1d)
  metab <- rnorm(n,5)
  df <- data.frame(t1d = t1d,metab = metab,methyl = methyl)
  return(df)
})
str_19_res <- sim_eval(dfs,correct,19)
```

Out of `r nsim` simulations, bnlearn picked the correct structure `r sum(str_19_res$correct)` times (`r sum(str_19_res$correct)/nsim * 100`%).

```{r,results='asis'}
invisible(table_eval(str_19_res,correct))
```

```{r}
plot_eval(str_19_res)
```

# Structure 20

```{r}
correct = "[methyl][t1d][metab|t1d]"
correct = modelstring(model2network(correct))
# Simulate
dfs <- lapply(1:nsim,function(x){
  set.seed(1016 + x)
  # Simulate data
  t1d <- rbinom(n,1,0.5)
  metab <- rnorm(n,mean = alpha*t1d)
  methyl <- rnorm(n,5)
  df <- data.frame(t1d = t1d,metab = metab,methyl = methyl)
  return(df)
})
str_20_res <- sim_eval(dfs,correct,20)
```

Out of `r nsim` simulations, bnlearn picked the correct structure `r sum(str_20_res$correct)` times (`r sum(str_20_res$correct)/nsim * 100`%).

```{r,results='asis'}
invisible(table_eval(str_20_res,correct))
```

```{r}
plot_eval(str_20_res)
```

# Structure 21

```{r}
correct = "[methyl][t1d|methyl][metab]"
correct = modelstring(model2network(correct))
# Simulate
dfs <- lapply(1:nsim,function(x){
  set.seed(1016 + x)
  # Simulate data
  methyl <- rnorm(n,5)
  t1d <- rbinom(n,1,prob = plogis(beta*methyl))
  metab <- rnorm(n,5)
  df <- data.frame(t1d = t1d,metab = metab,methyl = methyl)
  return(df)
})
str_21_res <- sim_eval(dfs,correct,21)
```

Out of `r nsim` simulations, bnlearn picked the correct structure `r sum(str_21_res$correct)` times (`r sum(str_21_res$correct)/nsim * 100`%).

```{r,results='asis'}
invisible(table_eval(str_21_res,correct))
```

```{r}
plot_eval(str_21_res)
```

# Structure 22

```{r}
correct = "[methyl][t1d][metab|methyl]"
correct = modelstring(model2network(correct))
# Simulate
dfs <- lapply(1:nsim,function(x){
  set.seed(1016 + x)
  # Simulate data
  methyl <- rnorm(n,5)
  t1d <- rbinom(n,1,0.5)
  metab <- rnorm(n,mean = gamma*methyl)
  df <- data.frame(t1d = t1d,metab = metab,methyl = methyl)
  return(df)
})
str_22_res <- sim_eval(dfs,correct,22)
```

Out of `r nsim` simulations, bnlearn picked the correct structure `r sum(str_22_res$correct)` times (`r sum(str_22_res$correct)/nsim * 100`%).

```{r,results='asis'}
invisible(table_eval(str_22_res,correct))
```

```{r}
plot_eval(str_22_res)
```

# Structure 23

```{r}
correct = "[metab][methyl][t1d|metab]"
correct = modelstring(model2network(correct))
# Simulate
dfs <- lapply(1:nsim,function(x){
  set.seed(1016 + x)
  # Simulate data
  metab <- rnorm(n,5)
  t1d <- rbinom(n,1,prob = plogis(alpha*metab))
  methyl <- rnorm(n,5)
  df <- data.frame(t1d = t1d,metab = metab,methyl = methyl)
  return(df)
})
str_23_res <- sim_eval(dfs,correct,23)
```

Out of `r nsim` simulations, bnlearn picked the correct structure `r sum(str_23_res$correct)` times (`r sum(str_23_res$correct)/nsim * 100`%).

```{r,results='asis'}
invisible(table_eval(str_23_res,correct))
```

```{r}
plot_eval(str_23_res)
```

# Structure 24

```{r}
correct = "[metab][methyl|metab][t1d]"
correct = modelstring(model2network(correct))
# Simulate
dfs <- lapply(1:nsim,function(x){
  set.seed(1016 + x)
  # Simulate data
  metab <- rnorm(n,5)
  t1d <- rbinom(n,1,0.5)
  methyl <- rnorm(n,mean = gamma*metab)
  df <- data.frame(t1d = t1d,metab = metab,methyl = methyl)
  return(df)
})
str_24_res <- sim_eval(dfs,correct,24)
```

Out of `r nsim` simulations, bnlearn picked the correct structure `r sum(str_24_res$correct)` times (`r sum(str_24_res$correct)/nsim * 100`%).

```{r,results='asis'}
invisible(table_eval(str_24_res,correct))
```

```{r}
plot_eval(str_24_res)
```

# Continuous Data

```{r}
correct = "[metab][methyl|metab][t1d|metab:methyl]"
correct = modelstring(model2network(correct))
# Simulate
dfs <- lapply(1:nsim,function(x){
  set.seed(1016 + x)
  # Simulate data
  metab <- rnorm(n,5)
  methyl <- rnorm(n, mean = gamma*metab)
  t1d <- rnorm(n,mean = alpha*metab + beta*methyl)
  df <- data.frame(t1d = t1d,metab = metab,methyl = methyl)
  return(df)
})
cont_res <- sim_eval(dfs,correct,25)
```

Out of `r nsim` simulations, bnlearn picked the correct structure `r sum(cont_res$correct)` times (`r sum(cont_res$correct)/nsim * 100`%).

```{r,results='asis'}
invisible(table_eval(cont_res,correct))
```

```{r}
plot_eval(cont_res)
```