---
title: "Triad Selection"
author: "Tim Vigers"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(arsenal)
library(skimr)
library(knitr)
library(tidyverse)
```

```{r import step 3,echo=FALSE}
methyl_step2 <- 
  read.csv("/Users/timvigers/Documents/GitHub/MS-Thesis/candidate_selection/step_2/methyl_adj.csv",stringsAsFactors = F)
methyl_step2$p.adj <- format(methyl_step2$p.value,digits = 3)
```

## Step 1: Metabolite - Methylation

For this step I ran unadjusted linear models. In order to narrow down the number of models, I only looked at the metabolites that have come up repeatedly in Liz's analyses. Metabolite and methylation data were converted to z scores prior to fitting the models.

<!-- ## GC-TOF Candidates -->

<!-- ```{r echo=FALSE} -->
<!-- candidates <- read.csv("/Users/timvigers/Documents/GitHub/MS-Thesis/data/metabolomics/liz_candidates.csv",stringsAsFactors = F) -->
<!-- gctof_candidates <- unique(candidates$gctof) -->
<!-- gctof_candidates <- gctof_candidates[order(as.numeric(sub(".*_","",gctof_candidates)))] -->
<!-- kable(gctof_candidates,row.names = F,col.names = "") -->
<!-- ``` -->

<!-- ### Unadjusted Model -->

<!-- ```{r eval=FALSE} -->
<!-- lm(cg00000029 ~ gctof_89, data = data) -->
<!-- ``` -->

<!-- #### Plots -->

<!-- ```{r gctof unadj import,echo=FALSE,cache=TRUE} -->
<!-- gctof <- read.csv("/Users/timvigers/Documents/GitHub/MS-Thesis/candidate_selection/step_1/sv/gctof_SV_unadj_scaled.csv",stringsAsFactors = F) -->
<!-- gctof <- gctof %>% arrange(p.value,Value) -->
<!-- hist(gctof$Value,xlab = expression(paste(beta," Value")),  -->
<!--      main = expression(paste("Histogram of ",beta," Values"))) -->
<!-- hist(gctof$p.value,xlab = "P Value", main = "Histogram of P Values") -->
<!-- ``` -->

<!-- ```{r echo=FALSE,cache=TRUE} -->
<!-- volcano <- ggplot(gctof,aes(x = Value, y = -log(p.value,base = 10))) + -->
<!--   geom_point() + theme_bw() + ylab("-log10 p-value") + xlab("Estimate") -->
<!-- volcano -->
<!-- ``` -->

<!-- #### Number of Candidates by Cutoff -->

<!-- For this, I took a list of methylation-metabolite pairs significant at a given cutoff (columns), and counted how many of those methylation sites were significantly associated with T1D at another range of p value cutoffs (rows). Models of the association between methylation and T1D were adjusted for age and sex. -->

<!-- ```{r echo=FALSE,cache=TRUE} -->
<!-- # Adjust p values -->
<!-- gctof$p.adj <- p.adjust(gctof$p.value, method = "fdr") -->
<!-- # Count number of pairs that are associated with methylation correlated with T1D -->
<!-- num_cand <- data.frame() -->
<!-- # Methylation < 0.05 -->
<!-- num_cand[1,"FDR < 0.05"] <- sum(unique(gctof$methyl[gctof$p.adj < 0.05]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.05])) -->
<!-- num_cand[1,"p < 1e-2"] <- sum(unique(gctof$methyl[gctof$p.value < 1e-2]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.05])) -->
<!-- num_cand[1,"p < 1e-3"] <- sum(unique(gctof$methyl[gctof$p.value < 1e-3]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.05])) -->
<!-- num_cand[1,"p < 1e-4"] <- sum(unique(gctof$methyl[gctof$p.value < 1e-4]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.05])) -->
<!-- # Methylation < 0.01 -->
<!-- num_cand[2,"FDR < 0.05"] <- sum(unique(gctof$methyl[gctof$p.adj < 0.05]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.01])) -->
<!-- num_cand[2,"p < 1e-2"] <- sum(unique(gctof$methyl[gctof$p.value < 1e-2]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.01])) -->
<!-- num_cand[2,"p < 1e-3"] <- sum(unique(gctof$methyl[gctof$p.value < 1e-3]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.01])) -->
<!-- num_cand[2,"p < 1e-4"] <- sum(unique(gctof$methyl[gctof$p.value < 1e-4]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.01])) -->
<!-- # Methylation < 0.001 -->
<!-- num_cand[3,"FDR < 0.05"] <- sum(unique(gctof$methyl[gctof$p.adj < 0.05]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.001])) -->
<!-- num_cand[3,"p < 1e-2"] <- sum(unique(gctof$methyl[gctof$p.value < 1e-2]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.001])) -->
<!-- num_cand[3,"p < 1e-3"] <- sum(unique(gctof$methyl[gctof$p.value < 1e-3]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.001])) -->
<!-- num_cand[3,"p < 1e-4"] <- sum(unique(gctof$methyl[gctof$p.value < 1e-4]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.001])) -->
<!-- # Row and column names -->
<!-- rownames(num_cand) <- c("< 0.05","< 0.01","< 0.001") -->
<!-- kable(num_cand) -->
<!-- ``` -->

<!-- #### Top 10 (Sorted by P Value Then Estimate)  -->

<!-- ```{r echo=FALSE} -->
<!-- gctof$p.value <- format(gctof$p.value,digits = 3) -->
<!-- gctof$p.adj <- format(gctof$p.adj,digits = 3) -->
<!-- kable(head(gctof,n=10),row.names = F) -->
<!-- ``` -->

## HILIC Candidates

```{r echo=FALSE}
hilic_candidates <- unique(candidates$hilic)
hilic_candidates <- hilic_candidates[order(as.numeric(sub(".*_","",hilic_candidates)))]
kable(hilic_candidates,row.names = F,col.names = "")
```

### Unadjusted Model

```{r eval=FALSE}
lm(cg00000029 ~ hilic_89, data = data)
```

#### Plots

```{r hilic unadj import,echo=FALSE,cache=TRUE}
hilic <- read.csv("/Users/timvigers/Documents/GitHub/MS-Thesis/candidate_selection/step_1/sv/hilic_SV_unadj_scaled.csv",stringsAsFactors = F)
hilic <- hilic %>% arrange(p.value,Value)
hist(hilic$Value,xlab = expression(paste(beta," Value")), 
     main = expression(paste("Histogram of ",beta," Values")))
hist(hilic$p.value,xlab = "P Value", main = "Histogram of P Values")
```

```{r echo=FALSE,cache=TRUE}
volcano <- ggplot(hilic,aes(x = Value, y = -log(p.value,base = 10))) +
  geom_point() + theme_bw() + ylab("-log10 p-value") + xlab("Estimate")
volcano
```

#### Number of Candidates by Cutoff

```{r echo=FALSE,cache=TRUE}
# Adjust p values
hilic$p.adj <- p.adjust(hilic$p.value, method = "fdr")
# Count number of pairs that are associated with methylation correlated with T1D
num_cand <- data.frame()
# Methylation < 0.05
num_cand[1,"FDR < 0.05"] <- sum(unique(hilic$methyl[hilic$p.adj < 0.05]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.05]))
num_cand[1,"p < 1e-2"] <- sum(unique(hilic$methyl[hilic$p.value < 1e-2]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.05]))
num_cand[1,"p < 1e-3"] <- sum(unique(hilic$methyl[hilic$p.value < 1e-3]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.05]))
num_cand[1,"p < 1e-4"] <- sum(unique(hilic$methyl[hilic$p.value < 1e-4]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.05]))
# Methylation < 0.01
num_cand[2,"FDR < 0.05"] <- sum(unique(hilic$methyl[hilic$p.adj < 0.05]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.01]))
num_cand[2,"p < 1e-2"] <- sum(unique(hilic$methyl[hilic$p.value < 1e-2]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.01]))
num_cand[2,"p < 1e-3"] <- sum(unique(hilic$methyl[hilic$p.value < 1e-3]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.01]))
num_cand[2,"p < 1e-4"] <- sum(unique(hilic$methyl[hilic$p.value < 1e-4]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.01]))
# Methylation < 0.001
num_cand[3,"FDR < 0.05"] <- sum(unique(hilic$methyl[hilic$p.adj < 0.05]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.001]))
num_cand[3,"p < 1e-2"] <- sum(unique(hilic$methyl[hilic$p.value < 1e-2]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.001]))
num_cand[3,"p < 1e-3"] <- sum(unique(hilic$methyl[hilic$p.value < 1e-3]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.001]))
num_cand[3,"p < 1e-4"] <- sum(unique(hilic$methyl[hilic$p.value < 1e-4]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.001]))
# Row and column names
rownames(num_cand) <- c("< 0.05","< 0.01","< 0.001")
kable(num_cand)
```

#### Top 10 (Sorted by P Value Then Estimate) 

```{r echo=FALSE}
hilic$p.value <- format(hilic$p.value,digits = 3)
hilic$p.adj <- format(hilic$p.adj,digits = 3)
kable(head(hilic,n=10),row.names = F)
```

## Lipid Candidates

```{r echo=FALSE}
lipid_candidates <- unique(candidates$lipid)
lipid_candidates <- 
  lipid_candidates[order(as.numeric(sub(".*_","",lipid_candidates)))]
kable(lipid_candidates,row.names = F,col.names = "")
```

### Unadjusted Model

```{r eval=FALSE}
lm(cg00000029 ~ lipid_79, data = data)
```

#### Plots

```{r lipid unadj import,echo=FALSE,cache=TRUE}
lipid <- 
  read.csv("/Users/timvigers/Documents/GitHub/MS-Thesis/candidate_selection/step_1/sv/lipid_SV_unadj_scaled.csv",
           stringsAsFactors = F)
lipid <- lipid %>% arrange(p.value,Value)
hist(lipid$Value,xlab = expression(paste(beta," Value")), 
     main = expression(paste("Histogram of ",beta," Values")))
hist(lipid$p.value,xlab = "P Value", main = "Histogram of P Values")
```

```{r echo=FALSE,cache=TRUE}
volcano <- ggplot(lipid,aes(x = Value, y = -log(p.value,base = 10))) +
  geom_point() + theme_bw() + ylab("-log10 p-value") + xlab("Estimate")
volcano
```

#### Number of Candidates by Cutoff

```{r echo=FALSE,cache=TRUE}
# Adjust p values
lipid$p.adj <- p.adjust(lipid$p.value, method = "fdr")
# Count number of pairs that are associated with methylation correlated with T1D
num_cand <- data.frame()
# Methylation < 0.05
num_cand[1,"FDR < 0.05"] <- sum(unique(lipid$methyl[lipid$p.adj < 0.05]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.05]))
num_cand[1,"p < 1e-2"] <- sum(unique(lipid$methyl[lipid$p.value < 1e-2]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.05]))
num_cand[1,"p < 1e-3"] <- sum(unique(lipid$methyl[lipid$p.value < 1e-3]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.05]))
num_cand[1,"p < 1e-4"] <- sum(unique(lipid$methyl[lipid$p.value < 1e-4]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.05]))
# Methylation < 0.01
num_cand[2,"FDR < 0.05"] <- sum(unique(lipid$methyl[lipid$p.adj < 0.05]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.01]))
num_cand[2,"p < 1e-2"] <- sum(unique(lipid$methyl[lipid$p.value < 1e-2]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.01]))
num_cand[2,"p < 1e-3"] <- sum(unique(lipid$methyl[lipid$p.value < 1e-3]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.01]))
num_cand[2,"p < 1e-4"] <- sum(unique(lipid$methyl[lipid$p.value < 1e-4]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.01]))
# Methylation < 0.001
num_cand[3,"FDR < 0.05"] <- sum(unique(lipid$methyl[lipid$p.adj < 0.05]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.001]))
num_cand[3,"p < 1e-2"] <- sum(unique(lipid$methyl[lipid$p.value < 1e-2]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.001]))
num_cand[3,"p < 1e-3"] <- sum(unique(lipid$methyl[lipid$p.value < 1e-3]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.001]))
num_cand[3,"p < 1e-4"] <- sum(unique(lipid$methyl[lipid$p.value < 1e-4]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.001]))
# Row and column names
rownames(num_cand) <- c("< 0.05","< 0.01","< 0.001")
kable(num_cand)
```

#### Top 10 (Sorted by P Value Then Estimate) 

```{r echo=FALSE}
lipid$p.value <- format(lipid$p.value,digits = 3)
lipid$p.adj <- format(lipid$p.adj,digits = 3)
kable(head(lipid,n=10),row.names = F)
```

## Oxylipin Candidates

### Unadjusted Model

```{r eval=FALSE}
lm(cg00000029 ~ oxylipin_1, data = data)
```

#### Plots

```{r oxylipin unadj import,echo=FALSE,cache=TRUE}
oxylipin <- read.csv("/Users/timvigers/Documents/GitHub/MS-Thesis/candidate_selection/step_1/sv/oxylipin_SV_unadj_scaled.csv",stringsAsFactors = F)
oxylipin <- oxylipin %>% arrange(p.value,Value)
hist(oxylipin$Value,xlab = expression(paste(beta," Value")),
     main = expression(paste("Histogram of ",beta," Values")))
hist(oxylipin$p.value,xlab = "P Value", main = "Histogram of P Values")
```

```{r echo=FALSE,cache=TRUE}
volcano <- ggplot(oxylipin,aes(x = Value, y = -log(p.value,base = 10))) +
  geom_point() + theme_bw() + ylab("-log10 p-value") + xlab("Estimate")
volcano
```

#### Number of Candidates by Cutoff

```{r echo=FALSE,cache=TRUE}
# Adjust p values
oxylipin$p.adj <- p.adjust(oxylipin$p.value, method = "fdr")
# Count number of pairs that are associated with methylation correlated with T1D
num_cand <- data.frame()
# Methylation < 0.05
num_cand[1,"FDR < 0.05"] <- sum(unique(oxylipin$methyl[oxylipin$p.adj < 0.05]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.05]))
num_cand[1,"p < 1e-2"] <- sum(unique(oxylipin$methyl[oxylipin$p.value < 1e-2]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.05]))
num_cand[1,"p < 1e-3"] <- sum(unique(oxylipin$methyl[oxylipin$p.value < 1e-3]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.05]))
num_cand[1,"p < 1e-4"] <- sum(unique(oxylipin$methyl[oxylipin$p.value < 1e-4]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.05]))
# Methylation < 0.01
num_cand[2,"FDR < 0.05"] <- sum(unique(oxylipin$methyl[oxylipin$p.adj < 0.05]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.01]))
num_cand[2,"p < 1e-2"] <- sum(unique(oxylipin$methyl[oxylipin$p.value < 1e-2]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.01]))
num_cand[2,"p < 1e-3"] <- sum(unique(oxylipin$methyl[oxylipin$p.value < 1e-3]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.01]))
num_cand[2,"p < 1e-4"] <- sum(unique(oxylipin$methyl[oxylipin$p.value < 1e-4]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.01]))
# Methylation < 0.001
num_cand[3,"FDR < 0.05"] <- sum(unique(oxylipin$methyl[oxylipin$p.adj < 0.05]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.001]))
num_cand[3,"p < 1e-2"] <- sum(unique(oxylipin$methyl[oxylipin$p.value < 1e-2]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.001]))
num_cand[3,"p < 1e-3"] <- sum(unique(oxylipin$methyl[oxylipin$p.value < 1e-3]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.001]))
num_cand[3,"p < 1e-4"] <- sum(unique(oxylipin$methyl[oxylipin$p.value < 1e-4]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.001]))
# Row and column names
rownames(num_cand) <- c("< 0.05","< 0.01","< 0.001")
kable(num_cand)
```

#### Top 10 (Sorted by P Value Then Estimate)

```{r echo=FALSE}
oxylipin$p.value <- format(oxylipin$p.value,digits = 3)
oxylipin$p.adj <- format(oxylipin$p.adj,digits = 3)
kable(head(oxylipin,n=10),row.names = F)
```

## Vitamin D Candidates

### Unadjusted Model

```{r eval=FALSE}
lm(cg00000029 ~ vitd_1, data = data)
```

#### Plots

```{r vitd unadj import,echo=FALSE,cache=TRUE}
vitd <- read.csv("/Users/timvigers/Documents/GitHub/MS-Thesis/candidate_selection/step_1/sv/vitd_SV_unadj_scaled.csv",stringsAsFactors = F)
vitd <- vitd %>% arrange(p.value,Value)
hist(vitd$Value,xlab = expression(paste(beta," Value")),
     main = expression(paste("Histogram of ",beta," Values")))
hist(vitd$p.value,xlab = "P Value", main = "Histogram of P Values")
```

```{r echo=FALSE,cache=TRUE}
volcano <- ggplot(vitd,aes(x = Value, y = -log(p.value,base = 10))) +
  geom_point() + theme_bw() + ylab("-log10 p-value") + xlab("Estimate")
volcano
```

#### Number of Candidates by Cutoff

```{r echo=FALSE,cache=TRUE}
# Adjust p values
vitd$p.adj <- p.adjust(vitd$p.value, method = "fdr")
# Count number of pairs that are associated with methylation correlated with T1D
num_cand <- data.frame()
# Methylation < 0.05
num_cand[1,"FDR < 0.05"] <- sum(unique(vitd$methyl[vitd$p.adj < 0.05]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.05]))
num_cand[1,"p < 1e-2"] <- sum(unique(vitd$methyl[vitd$p.value < 1e-2]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.05]))
num_cand[1,"p < 1e-3"] <- sum(unique(vitd$methyl[vitd$p.value < 1e-3]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.05]))
num_cand[1,"p < 1e-4"] <- sum(unique(vitd$methyl[vitd$p.value < 1e-4]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.05]))
# Methylation < 0.01
num_cand[2,"FDR < 0.05"] <- sum(unique(vitd$methyl[vitd$p.adj < 0.05]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.01]))
num_cand[2,"p < 1e-2"] <- sum(unique(vitd$methyl[vitd$p.value < 1e-2]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.01]))
num_cand[2,"p < 1e-3"] <- sum(unique(vitd$methyl[vitd$p.value < 1e-3]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.01]))
num_cand[2,"p < 1e-4"] <- sum(unique(vitd$methyl[vitd$p.value < 1e-4]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.01]))
# Methylation < 0.001
num_cand[3,"FDR < 0.05"] <- sum(unique(vitd$methyl[vitd$p.adj < 0.05]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.001]))
num_cand[3,"p < 1e-2"] <- sum(unique(vitd$methyl[vitd$p.value < 1e-2]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.001]))
num_cand[3,"p < 1e-3"] <- sum(unique(vitd$methyl[vitd$p.value < 1e-3]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.001]))
num_cand[3,"p < 1e-4"] <- sum(unique(vitd$methyl[vitd$p.value < 1e-4]) %in% unique(methyl_step2$methyl[methyl_step2$p.value < 0.001]))
# Row and column names
rownames(num_cand) <- c("< 0.05","< 0.01","< 0.001")
kable(num_cand)
```

#### Top 10 (Sorted by P Value Then Estimate)

```{r echo=FALSE}
vitd$p.value <- format(vitd$p.value,digits = 3)
vitd$p.adj <- format(vitd$p.adj,digits = 3)
kable(head(vitd,n=10),row.names = F)
```