---
title: "MS Thesis Work"
author: "Tim Vigers"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = "/Users/timvigers/GitHub/MS-Thesis")
library(mediation)
library(parallel)
library(flextable)
library(diagram)
library(tidyverse)
set.seed(1017)
```

# CIT Results

Of the metabolite - methylation pairs significant by CIT, the there was a relatively even split in terms of which was the mediator:

```{r}
# Import cit results
load("./data/networks/cits.Rdata")
# Annotate
gctof_anno = read.csv("./data/metabolomics/gctof.featureAnno.csv") %>% 
  select(BinBase.name,feature_name) %>% rename(name = BinBase.name)
hilic_anno = read.csv("./data/metabolomics/hilic.featureAnno.csv") %>%
  select(Metabolite.name,feature_name) %>% rename(name = Metabolite.name)
lipid_anno = read.csv("./data/metabolomics/lipid.featureAnno.csv") %>%
  select(Metabolite.name,feature_name) %>% rename(name = Metabolite.name)
vitd_anno = read.csv("./data/metabolomics/vitD.featureAnno.csv") %>%
  select(originalName,boxcoxDat_name) %>% 
  rename(name = originalName,feature_name = boxcoxDat_name)
oxylipin_anno = 
  read.csv("./data/metabolomics/oxylipin.featureAnno.csv") %>%
  select(full.name,cleanedDat_name) %>% 
  rename(name = full.name,feature_name = cleanedDat_name)
oxylipin_anno$feature_name = paste0("bc_",oxylipin_anno$feature_name)

anno = bind_rows(gctof_anno,hilic_anno,lipid_anno,vitd_anno,oxylipin_anno)

cits$metabolite = anno$name[match(cits$metab,anno$feature_name)]
# Repeats
repeats = cits[,c("methyl","metabolite")]
repeats = repeats[duplicated(repeats),]
# Convert arrow
cits$Mediator = factor(cits$direction,levels = c("<",">"),
                       labels = c("Methylation","Metabolite"))
# Print
table(cits$Mediator) %>% as.data.frame(.) %>% rename(Mediator = Var1) %>%
  flextable()
```

Of the above `r nrow(cits)` pairs, `r nrow(repeats)` were significant with both the metabolite and methylation probe as a mediator:

```{r}
repeats %>% rename(Methylation = methyl,Metabolite = metabolite) %>%
  flextable() %>% set_table_properties(.,layout = "autofit",width = 0.75)
```

# Less conservative mediation analysis

```{r}
load("./data/mediation/cits_mediation.Rdata")
# Plot
data <- c(0, "'a'", 0,
          0, 0, 0, 
          "'b'", "'c'", 0)
M<- matrix (nrow=3, ncol=3, byrow = TRUE, data=data)
plot<- plotmat (M, pos=c(1,2), 
                name= c( "M","X", "T1D Status"), 
                box.type = "rect", box.size = 0.12, box.prop=0.5,  curve=0)
```

Total effect = c + ab, or the effect of X on Y without controlling for M.

c = the direct effect of X on Y after controlling for M 

Amount of mediation = ab = total effect - c

## Volcano plot of ACME

```{r}

```
